Rotating Stars from \Kepler Observed with Gaia DR2

### 2018-05-07 paper/ms.tex
outlined intro

\shorttitle{Rotating Stars}
\shortauthors{Davenport et al.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Astrometric data from the recent Gaia Data Release 2 has been matched against the sample of stars from \Kepler with known rotation periods. 
\end{abstract}

@@

The \Kepler mission \citep{borucki2010}

\citet{mcquillan2013} bimodal period distribution 
\citep{mcquillan2014}.

\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause. sub giant contamination was the thing. Now we can finally follow up on this and study the full \Kepler rotation period sample

 \citep{gaia} mission

In this letter

%%%%%%%%%%%%%%%%%%%%%%
\section{The \Kepler--Gaia Data}
%Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation. The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}. Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.

@@

\begin{figure}[]
\centering
\includegraphics[width=3.5in]{../figures/delta_per}
\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}.
}
\label{fig:per_hist}
\end{figure}

\begin{figure*}[]
\centering
\includegraphics[width=6.5in]{../figures/cmd_zoom}
\caption{
}
\label{fig:cmd_zoom}
\end{figure*}

>>>>

\shorttitle{Rotating Stars from \Kepler Observed with Gaia DR2}
\shortauthors{Davenport et al.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We have matched the astrometric data from {\em Gaia} Data Release 2 to the sample of stars with measured rotation periods from \Kepler. Using 30,305 stars with good distance estimates, we select 16,248 as being likely main sequence single stars within a 0.5 mag region about a 1 Gyr isochrone. This removes sub-giants and unresolved binary stars from the sample.
The rotation period bimodality, originally discovered by \citet{mcquillan2013}, is recovered for stars out to 525pc, but is not detectable at further distances. 
We also find a significant width in the stellar main sequence of $M_G\sim$0.25 mag, as well as a  increase in the average rotation period correlated with the $M_G$ offset at a given color (mass). We interpret this to be the measurable change in luminosity and loss of angular momentum as stars evolve along the main sequence, which may provide a new independent test of stellar evolution and gyrochronlogy models.
This investigation represents the first step in understanding the star formation history of our solar neighborhood as traced through stellar angular momentum loss. 
\end{abstract}

@@

The \Kepler mission \citep{borucki2010} a new era for enabling detailed study of angular momentum loss in low-mass stars. long noted as a means to possibly age-date stars (gyro), open clusters give hope that this general model works for low-mass stars
many Qs exist about details. these include Qs about initial rotation period distribution (CITE), the specific prescription for spin-down (ANGUS, etc), and exploring the efficiency of this angular momentum loss mechanism at old ages (Van Saders).

one of the most compelling results from the rotation work in \Kepler is the discovery of a period bimodality. 
\citet{mcquillan2013} found bimodal feature, in M and later K stars \citep{mcquillan2014}. Davenport used Gaia DR1 to remove contamination from sub-giants and found the feature in G dwarfs. this feature proposed to be either a new short-lived transition or instability phase of momentum loss, or a signature of star formation history imprinted in the present-day rotation period distribution. However, to date this feature has only been observed in the \Kepler rotation period catalog, and most critically only for stars within $\sim$300 pc.

In this letter we follow the work of \citet{davenport2017} in studying the \Kepler rotation period sample using astrometric data from the {\em Gaia} mission \citep{gaia}. By matching the \citet{mcquillan2014} rotation period catalog to the newest data from {\em Gaia} Data Release 2, we can use precise distances for essentially every star to select likely main sequence dwarfs to distances well over 1 kpc. Importantly this filters out both sub-giants (the main contaminant noted by \cite{davenport2017}), and unresolved binary stars.
Here we demonstrate the power of such a combined time-domain and astrometric sample for constraining the detailed evolution of main sequence stars themselves, and exploring the star formation history of the Milky Way.

%\citet{davenport2017} looked at this with Gaia DR1, extending period gap to G dwarfs, pointing to SFH as cause. sub giant contamination was the thing. Now we can finally follow up on this and study the full \Kepler rotation period sample

%%%%%%%%%%%%%%%%%%%%%%
\section{The \Kepler--Gaia Data}

%Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation. The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}. Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.

@@

\begin{figure}[]
\centering
\includegraphics[width=3.5in]{../figures/delta_per}
\caption{Histograms of the log rotation periods after a 600 Myr gyrochrone was subtracted for stars in the same five distance bins shown in Figure \ref{fig:color_period}. The period bimodality for stars within 350 pc has two peaks similar to those found in \citet{davenport2017}, at -0.15 and +0.18 dex. The fast rotating peak (left side) declines sharply at further distances, however.
}
\label{fig:per_hist}
\end{figure}

\begin{figure*}[]
\centering
\includegraphics[width=6.5in]{../figures/cmd_zoom}
\caption{Left: Enlarged portion of the color--magnitude diagram from Figure \ref{fig:cmd} in a region centered near $\sim$0.75 M$_\odot$, with MIST isochrones at ages of $10^8$, $10^9$, and $10^10$ yr for comparison (blue, black, and red lines). Points are colored by their measured \Kepler rotation periods from \citet{mcquillan2014}. The main sequence shows significant scatter, as well as a gradient in rotation period. Right: Difference in $M_G$ from the $10^9$ yr MIST isochrone, shown as a function of their rotation periods (point color again indicates rotation period). Blue squares show an increase in the median $M_G$ offset in bins of increasing rotation period. Error bars shown are the standard deviation of $\Delta M_G$ in each bin.
}
\label{fig:cmd_zoom}
\end{figure*}
